raptr: Representative and Adequate Prioritization Toolkit in R

### 2016-11-23 code/rmarkdown/article.Rmd
+ add in readmes + compile manuscript into single pdf and single docx

---
title: "raptr: Representative and Adequate Prioritisation Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^{1,3}$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | $^3$The Nature Conservancy, South Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, preserving the adaptive processes currently acting on a species might be crucial for its long-term persistence in a world where environmental change is accelerating. By conserving individuals with the ability to persist in a range of climatic conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Despite this, current approaches overwhelmingly focus on securing a target proportion of each features' geographic range and little attention has been directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribution. Additionally, users set attribute \"space targets\" to ensure that reserve systems also secure a representative sample of evolutionary or ecologically relevant attributes (space) across a feature's geographic distribution (eg. an attribute space expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compare them with solutions generated using both amount and attribute space targets.

@@

4. The \\texttt{raptr R} package provides a unified framework for achieving spatial conservation prioritisations that secure an adequate and representative sample of features. Prioritisations that achieve this are likely to result in a greater chance of achieving long-term biodiversity persistence than area-based planning alone.

@@

```{r, include=FALSE}
# set wd
knitr::opts_knit$set(root.dir=normalizePath('../..'))
```

```{r, include=FALSE}
## initialization
# load .rda file
checkpoint::checkpoint('2016-11-01', R.version='3.3.1', scanForPackages=FALSE)
session::restore.session('data/final/results.rda')

## preliminary processing
for (i in seq_len(nlayers(cs1.bioclim.RAST)))
  cs1.bioclim.RAST@layers[[i]]@file@name <- paste0(getwd(), strsplit(cs1.bioclim.RAST@layers[[i]]@file@name, 'raptr-manuscript')[[1]][[2]])

```

<!--  title page -->
Short running title: Representative and adequate prioritisations \newline
Abstract word count: XXX / XXX  
Total word count: XXX / XXX  
Number of references: XXX  
Number of figures: XXX  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: XXX  
\newpage

Perhaps the most fundamental aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. populations, species, ecosystems), but also crucially the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r484; @r440]. Reserve networks buffer species from gross threatening processes and set the stage for enhanced management interventions [@r465]. Since the resources available for conservation action are limited, protected area networks must be sited in places that satisfy conservation objectives for minimal cost [@r255].

Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Zonation} and \texttt{Marxan}; @r429; @r22]. Decision makers can use these tools to obtain solutions that secure a proportion of the geographic range of each biological feature (populations, species, or ecosystems) of conservation by setting targets. One method to incorporate data on the ecological and environmental processes that sustain biological features is to conserve a representative sample of the physical attributes that underpin these processes. To achieve this, current methods typically involve partitioning each feature into sub-groups based on an attribute variable that relates to a biodiversity process of interest [@r9; @r247; @r448]. For instance, by dividing species distributions into sub-groups according to habitat discontinuities, and ensuring that each sub-group is represented in the solution, conservation planners can obtain prioritisations that promote adaptive processes [@r2]. However, using this method is challenging because biodiversity data are often continuous and multi-dimensional.

Simply partitioning features based on continuous variables and setting targets for each sub-group may not necessarily result in a solution that secures a representative sample of the continuous variable. For instance, if a feature was split into four sub-features based on a single continuous variable, and a solution preserve 10 % of the geographic range of each sub-feature, how much of the original continuous variation does the solution represent? The answer is not immediately clear. If the solution sites reserves evenly along the original variable, then the solution may sample far more than 10 % of the original variation. Whereas, if the solution sites reserves near the values used to divide the continuous values [as observed when partitioning continuous geographic variation based on political boundaries; @r483], then it might sample much less of the original variation. Conventional reserve selection problems cannot differentiate between these solutions--they are both optimal--though the poorer solution may be favored when using additional constraints to reduce cost or fragmentation. This limitation has been known of quite some time. In fact, it dates back to some of the earliest reserve selection methods [eg. @r295].

During the early 1990s, Faith and Walker [-@r480] developed a decision support tool (\texttt{DIVERSITY}) to identify prioritisations that secure a representative sample of continuous environmental variation across a study area. Although this method was originally pitched as an alternative to species-based conservation planning [@r481; @r218; @r203], it can also identify solutions that conserve a representative sample of variation across a single species' range. Recent work has built on this reserve selection method by solving problems with more advanced optimisation algorithms [@r482]. However, \texttt{DIVERSITY} has limited utility for species-based conservation planning, in part, because it cannot accommodate multiple biological features. Additionally, unlike \texttt{Marxan} and \texttt{Zonation}, users cannot set intuitive targets to obtain cost effective solutions.

Conservation planners lack a decision support tool that lets them set explicit targets to obtain solutions that (1) secure an adequate amount of habitat for each feature and (2) a representative sample of each feature. To begin to fill this gap, here we unite the ideas underpinning \texttt{DIVERSITY} and \texttt{Marxan} into new formulations of the reserve selection problem, and implement them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations based on data that relate to biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore its functionality.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic variation (eg. genetic or trait) variation that is extrinsic to the feature (eg. environmental conditions). There should, however, be a reasonable underlying hypothesis that relates this variation to the biodiversity processes that the decision maker aims to preserve.

A set of attributes is conceptualised as a space (termed an "attribute space"). For example, a decision maker may require a prioritisation that represents populations along climatic gradients. To achieve this, the decision maker might use a "climatic" attribute space with dimensions relating to mean annual temperature ($^{\circ}$C) and precipitation (mm). Any given combination of temperature and precipitation may be conceived as a point in this environmental space. Although they exist as polygons, for simplicity, each planning unit may be thought to exist as a single point inside a given space. By associating the planning units with climatic data, and calculating a representative statistic for each planning unit (eg. mean), they can be mapped from geographic space to this environmental attribute space. In addition to planning units, attribute spaces also contain demand points.

Demand points [@r480; @r218; @r203] are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space (see below for mathematical expressions). The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider the following example: we wish to develop a prioritisation for a single species that has four populations. However, we can only afford to preserve three of the four populations. To ensure that we preserve a representative sample, we want to preserve populations inhabiting different environmental conditions. We obtained annual rainfall (mm) and temperature ($^{\circ}$C) data at the location of each population. We used this data to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points inside this space. By comparing the distributions of demand points and populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that a solution that prioritises both populations _A_ and _C_, effectively "doubles-up" on the same environmental characteristics and constitute considerable redundancy if both selected. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$ (or $C$), $B$, and $D$. This example demonstrates how attribute spaces and demand points relate to biodiversity processes, and can be used to guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on elements of the \texttt{Marxan}, \texttt{DIVERSITY}, and uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are based on unreliable and reliable facility location problems and have been named accordingly [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied by features when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the set.

Define $F$ to be the set of features one wishes to conserve (indexed by $f$). Let $J$ be a set of planning units (indexed by $j$). Also, let $A_j$ denote the area, and $C_j$ denote the cost of preserving planning unit $j \in J$. To assess the extent to which each feature is secured in a given prioritisation, let $q_{fj}$ denote the probability of feature $f$ occupying planning unit $j$. The level of fragmentation associated with a prioritisation is parametrised as the total exposed boundary length (as in \texttt{Marxan}). Let the shared edges between each planning unit $j \in J$ and $k \in J$ be $e_{jk}$.

Let $S$ denote a set of attribute spaces (indexed by $s$). Each $j \in J$ is associated with spatially explicit data that represent coordinates for each attribute space $s \in S$. Let $I_{fsi}$ denote a set of demand points (indexed by $i$) for each feature $f \in F$ and each attribute space $s \in S$. Let $\lambda_{fsi}$ denote the weighting for each demand point $i \in I$, $f \in F$ and $s \in S$. Let $d_{fsij}$ denote the distance between each demand point $i \in I$ and each planning unit $j \in J$ for each feature $f \in F$ and attribute space $s \in S$. To describe the inherent variation in the distribution of demand points for feature $f$ and space $s$, let $\delta_{fsi}$ denote the distance between each demand point $i \in I$ and the centroid of the demand points. Demand points with greater weight $\lambda_{fsi}$ are more important, and the solutions that select planning units close to highly weighted demand points will be more favorable. As a consequence, the decision maker will need to choose an appropriate weighting for each demand point.

Targets are used to ensure that prioritisations adequately preserve each feature. Amount-based targets are used to ensure that the total amount of habitat preserved is sufficient. Let $T_f$ denote the expected amount of area that needs to be preserved for each feature $f \in F$. Space-based targets ensure that a representative sample of each feature is secured. Let $\tau_{fs}$ denote the space-based targets for feature $f \in F$ and attribute space $s \in S$. For convenience, all targets are expressed as proportions in the \texttt{R} package.

The control variables for the unreliable formulation are the $B$, $T_{s}$, and $\tau_{fs}$ variables.

\begin{align*}
T_s &= \text{amount target for feature $f$} \tag*{eqn 1a}\\
%
\tau_{fs} &= \parbox{25em}{representation target for feature $f$ in attribute space $s$ according to the corresponding demand points} \tag*{eqn 1b}\\
%
B &= \text{boundary length modifier (BLM) to penalise fragmented solutions} \tag*{eqn 1c}\\
\end{align*}

The decision variables are the $X_j$ and $Y_{fsij}$ variables.

\begin{align*}
X_j
  &= \begin{cases}
    1, & \parbox{25em}{if planning unit $j$ is selected for conservation action} \tag*{eqn 2a} \\
    0, & \parbox{25em}{otherwise} \\
  \end{cases} \\
%
Y_{fsij} &= \begin{cases}
    1, & \parbox{25em}{if demand point $i$ is used to represent planning unit $j$ for feature $f$ in space $s$. } \tag*{eqn 2b} \\
    0, & \parbox{25em}{otherwise} \\
  \end{cases} \\
\end{align*}

Each demand point $i \in I$ for feature $f \in F$ and space $s \in S$ is assigned to a selected planning unit $J$ where $X_j = 1$. The weighted distance between the demand point and its assigned planning unit $\lambda_{fsi} d_{fsij}$ is used to assess how well the demand point is represented in a given solution. Generally, demand points are represented by their closest selected planning units (unless particularly low space-based targets are used).

The unreliable formulation (URAP) is a defined as a multi-objective optimisation problem.

\begin{align*}
& \text{(URAP)} & \text{Min } & \sum_{j=0}^{J-1} \left( X_j C_j \right) + \sum_{j=0}^{J-1} \sum_{k=j}^{J-1} X_j \left( 1-X_k \right) \left( B e_{jk} \right) + & & \tag*{eqn 3a} \\
%
& & \text{s.t. } & \sum_{j=0}^{J-1} A_j q_{fj} \geq T_{f} & \forall & 0 \leq f \leq F-1 \tag*{eqn 3b}\\
%
& & & 1 - \frac{\sum_{i=0}^{I-1} \sum_{j=0}^{J-1} \lambda_{fsi} {d_{fsij}}^{2} Y_{fsij}}{\sum_{i=0}^{I-1} \lambda_{fsi} {\delta_{fsi}}^{2}} \geq \tau_{fs} & \forall & 0 \leq f \leq F-1, \tag*{eqn 3c}\\
& & & & & 0 \leq s \leq S-1\\
%
& & & \sum_{j=0}^{J-1} Y_{fsij} = 1 & \forall & 0 \leq f \leq F-1, \tag*{eqn 3d}\\
& & & & & 0 \leq s \leq S-1, \\
& & & & & 0 \leq i \leq I-1\\
%
& & & Y_{fsij} \leq X_j & \forall & 0 \leq f \leq F-1, \tag*{eqn 3e}\\
& & & & & 0 \leq s \leq S-1, \\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J-1\\
%
& & & X_j, Y_{fsij} \in {0,1} & \forall & 0 \leq f \leq F-1, \tag*{eqn 3f}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1\\
%
\end{align*}

The objective function (eqn 3a) determines the utility of a given prioritisation: a combination of the total cost of a prioritisation and how fragmented it is. Constraints (eqn 3b) ensure that all the amount-based targets are met. Constraints (eqn 3c) ensure that all the space-based targets are met for each feature and each attribute space. For each feature and attribute space, the total weighted distance between the demand points and their closest selected planning units is calculated ($\lambda_{fsi} d_{fsij} Y_{fsij}$). This total weighted distance is then scaled by the inherent variation in the demand points ($\lambda_{fsi} \delta_{fsi}$). The resulting fraction yields a proportion conceptually similar to the $R^2$ statistic used in $k$-means clustering analysis. The constraints ensure that this proportion must be equal to or greater than the space-based target. Constraints (eqn 3d) ensure that only one planning unit is assigned to each demand point. Constraints (eqn 3e) ensure that demand points are only assigned to selected planning units. Constraints (eqn 3f) ensure that the $X$ and $Y$ variables are binary.

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithms. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Appendix S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

# Examples

To understand the behaviour of the unreliable problem and showcase its value, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritisations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can gauge the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with planning units arranged in a $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Fig. 2). We then simulated three species across this study area. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Fig. 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Fig. 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8c). The third species represents a species with two distinct populations (Fig. 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8d). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8d. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

\begin{align*}
\\
P \left( \text{uniformly distributed species} | \left( x, y \right) \right) &= 0.1 \tag*{eqn 8a} \\
%
f \left(z, \mu, \Sigma \right) &= \left(2 \pi \right) ^{-1} | \Sigma | e ^{- \frac{1}{2} \left(z - \mu \right)' \Sigma ^{-1} \left(z - \mu \right)} \tag*{eqn 8b} \\
%
P \left( \text{normally distributed species} | \left( x , y \right) \right) &= \frac{f \left( \left[ \begin{smallmatrix} x \\ y \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} 0.0 \\ 0.0 \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} 12.58&0 \\ 0&12.5 \end{smallmatrix} \right] \right)}{2} \tag*{eqn 8c} \\
%
P \left( \text{bimodally distributed species} | \left( x,y \right) \right) &= \text{Max} \begin{cases}
f \left(\left[ \begin{smallmatrix} x \\ y \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} -3.75 \\ -3.75 \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} 10&0 \\ 0&10 \end{smallmatrix} \right] \right), \\ \frac{f \left( \left[ \begin{smallmatrix} x \\ y \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} 3.75 \\ 3.75 \end{smallmatrix} \right] \, \left[ \begin{smallmatrix} 8&0 \\ 0&8 \end{smallmatrix} \right] \right)} {2} \\
\end{cases} \tag*{eqn 8d} \\
%
\end{align*}

We generated four solutions for each species. First to represent solutions generated using conventional conservation planning methods, we generated solutions using only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Second to show how the addition of geographic targets can affect a prioritisation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets. Third to represent solutions using conventional planning methods that penalise for fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using geographic targets and penalising fragmentation, we generated solutions using using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

### Uniformly distributed species

The uniform species was simulated to occur with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This specific configuration is an artifact of the \texttt{Gurobi} solver used to solve the problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). Note that the proportion here is negative because it performs worse than simply siting a planning unit in the centroid of the species' demand points.

The addition of a `r sim.params.LST[[MODE]]$space.target*100` % geographic target resulted in a solution that secured a representative sample of the uniform species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Since all planning units have equal cost and an equal chance of being occupied, this solution has the same number of planning units as the solution generated using only amount targets (cf. Fig 3a). Although the use of amount and space based targets has addressed adequacy and representativeness objectives (respectively), they have resulted in a highly fragmented solution.

The addition of a positive boundary length modifier (BLM) parameter resulted in a well-connected solution that secured an adequate proportion (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$amount.held,2)` % ) and representative sample (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$space.held,2)` % sampled) of the species geographic distribution (Fig. 3d). This solution contains `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units--a few more than the previously discussed solutions--to ensure that the solution secures a representative proportion of the species geographic distribution in a configuration that is not highly fragmented. This result suggests that the combination of space targets and boundary length modifiers may yield solutions that contain more planning units, since the solution generated using amount targets and a boundary length modifier (Fig. 3c) contained the same number of planning units as the solution with just the amount targets (Fig. 3a). Overall, these results show that under the simplest of conditions, the reliance on just amount targets can cause an under-specified reserve selection problem that is unlikely to return a suitable solution for implementation.

### Normally distributed species

The normally distributed species was simulated to contain a single core area where individuals were most prevalent, and marginal areas where individuals were less likely to occur (Fig. 2b). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets contained `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount\ntargets')$n` planning units and concentrated conservation efforts in the core area (Fig. 3e). Since all planning units have equal costs and areas, this solution contains the planning units with the highest probabilities of occupancy. Whilst this solution satisfies the adequacy objective for a prioritisation in a cost-effective manner--it fails to fulfill the representative objective for a prioritisation (`r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % distribution sampled).

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets resulted in a solution that is both adequate and representative in terms of the uniform species' distribution (Fig. 3f). This solution used `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$n` planning units to secure `r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$amount.held,2)` % and sample `r round(filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % of the normal species' distribution. By using amount and geographic targets in the reserve selection problem, we have obtained a solution that concentrates conservation effort in the range core and also the range margin. However, similar to the corresponding solution for the uniform species (Fig. 3b), this solution is highly fragmented.

By using `r sim.params.LST[[MODE]]$amount.target*100` % amount targets and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets and a boundary length modifier parameter ($BLM = 1$), we obtained a solution that fulfills adequacy, representativeness and connectivity objectives (Fig. 3h). This solution contains `r filter(sim.spp.DF, species=='Normal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units. Similar to the corresponding solution for the uniform species (Fig. 3d), this solution contains more planning units than any other solutions for this species (cf. Figs. 3e--3g). The results for the normally distributed species suggest that the space targets can result in solutions that secure a more representative sample of the species--even for species without significant spatial structure.

### Bimodally distributed species

The bimodally distributed species was simulated to represent a species with a highly structured distribution, and it contains two distinct populations (Fig. 2c). The solution generated using just `r sim.params.LST[[MODE]]$amount.target*100` % amount targets assigned conservation effort to just one of the two populations (Fig. 3i). While this solution secured an adequate proportion of the species' distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$amount.held,2)` % secured), it did not sample a representative proportion of the range (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled). The addition of boundary length modifiers to amount targets resulted in a solution which sampled even less of the species distribution (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount target\nand BLM')$space.held,2)` % sampled; Fig. 3k). However, the addition of geographic space targets resolved these issues.

The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount and `r sim.params.LST[[MODE]]$space.target*100` % geographic targets included planning units from both populations (Fig. 3j). Although this solution required more planning units to fulfill both targets ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$n`$), this solution secured a representative sample of the bimodally distributed species (`r round(filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % range sampled). Similar to corresponding solutions generated for both the uniformly (Fig. 3b) and normally distributed species (Fig. 3f), this solution was also highly fragmented and we could obtain a more well-connected solution at the expense of selecting additional planning units ($n=`r filter(sim.spp.DF, species=='Bimodal\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n`$; Fig. 3l). The results for the bimodally distributed species suggest that species with highly structured populations or strong variation between individuals could benefit the most from prioritisations that are generated using space-based targets.

## CASE STUDY 1
### Methods
We investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. By preserving the populations in suitable habitats with different environmental conditions, conservation planners can preserve the species' adaptive landscape and foster resilience against environmental change [@r63]. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subjected them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principal components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Fig. 4).

We selected four bird species that span a range of different evolutionary histories, distributions and ecologies: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), brown falcon (_Falco berigora_), and pale headed rosella (_Platycercus adscitus_). We then mapped the extent of occurrence for each species (Fig. 5). To do this, we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is entirely reproducible using freely available data.

We generated `r cs1.params.LST[[MODE]]$dp.number` demand points for each species (Fig. S2). We used the following method to ensure that the demand points reflected the core parts of the species' realised niches. For each species, we generated random points inside the species range and at each point extracted the principal component values at that location. We then fit hyperbox kernels to the distribution of principal component values to characterise the realised niche of each species [using a manually chosen bandwidth of `r cs1.params.LST[[MODE]]$dp.bandwidth[1]` and a `r cs1.params.LST[[MODE]]$dp.quantile` quantile to map the core parts of the species' niches; implemented in the \texttt{hypervolume R} package; @r231]. We then generated uniformly distributed points inside the species' kernels and estimated the density of the training points at the uniformly generated points. These uniformly distributed points and associated density estimated were used as demand point coordinates and weights (respectively).

We generated two solutions. The first solution was generated using `r cs1.params.LST[[MODE]]$amount.target*100` % amount-based targets for all species. The second solution was generated the same amount-based target with an additional `r cs1.params.LST[[MODE]]$space.target*100` % niche-based target for each species.

### Results
Generally, the solution generated using amount targets preserved a representative sample of each the `r as.character(as.english(length(cs1.params.LST[[MODE]]$species.names)))` bird species' niches--with `r ifelse(length(amount.not.represented.species.names)>1, as.character(as.english(length(amount.represented.species.names))), 'one')` excpetion. Specifically, this solution sampled over `r cs1.params.LST[[MODE]]$space.target*100` % of the realised niche of `r parsed.representative.space.held.names`. Yet it failed to achieve this for `r parsed.not.representative.space.held.names`. This result demonstrates that though solutions generated using conventional methods may preserve a representative sample of some features (Figure S1), only through the use of explicit targets can decision makers be sure that solutions secure a representative sample for all features.

>>>>

---
title: "raptr: Representative and Adequate Prioritization Toolkit in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^{1,3}$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | $^3$The Nature Conservancy, South Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimization", "protected area"]
abstract: "

1. An underlying aim in conservation planning is to maximize the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, preserving the adaptive processes currently acting on a species might be crucial for its long-term persistence in a world where environmental change is accelerating. By conserving individuals with the ability to persist in a range of climatic conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Despite this, current approaches overwhelmingly focus on securing a target proportion of each features' geographic range and little attention has been directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritization Toolkit in R} to guide reserve selection using targets for representing spatially explicit variables that underpin biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the features' geographic distributions. Additionally, users set attribute \"space targets\" to ensure that reserve systems also secure a representative sample of ecologically or evolutionary relevant attributes (space) across the features' geographic distributions (eg. an attribute space expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compare them with solutions generated using both amount and attribute space targets.

@@

4. The \\texttt{raptr R} package provides a unified framework for achieving spatial conservation prioritizations that secure an adequate and representative sample of features. Prioritizations that achieve this are likely to result in a greater chance of achieving long-term biodiversity persistence than area-based planning alone.

@@

Short running title: Representative and adequate prioritizations \newline
Abstract word count: XXX / XXX  
Total word count: XXX / XXX  
Number of references: XXX  
Number of figures: XXX  
Number of tables: 0  
Number of textboxes: 0  
Number of equations: XXX

\clearpage 
\input{./text.tex}

\clearpage 
\input{./figures.tex}

\clearpage 
\input{./supporting-information.tex}
